Catastrophic Goodhart: regularizing RLHF with KL divergence does not mitigate heavy-tailed reward misspecification

### 2024-10-24 paper/main.tex
save best-of-n figures and make various paper edits

Because sampling is inefficient at probing the extreme tail, we also find token sequences that optimize Starling 7B-alpha for reward. We considered Greedy Coordinate Gradient (GCG) from \citep{zou2023universal}, a method used to find adversarial suffixes that circumvent jailbreaking, but decided on a faster version of GCG called Accelerated Coordinate Gradient (ACG) from \citep{haizelabs2024acg}. See Table \ref{table1} for ACG hyperparameters.

@@

Possible distributions for V are normal and t-distribution with df=10.
Possible distributions for X are normal, t with df=3, t with df=5, lognormal, and Levy. All of these are heavy-tailed except for normal.
V is scaled to a standard deviation of 2 and X has s.d. of 1 (except for the Levy distribution, which has infinite variance), representing that in ordinary regimes most of the variance comes from utility rather than error.

\begin{figure}
    \centering
    \makebox[\textwidth][c]{\includegraphics[width=1.5\linewidth]{images/bon_synthetic.png}}
    \caption{When the error $X$ is normal and thus light-tailed, $V$ increases monotonically with $N$, consistent with our Theorem \ref{thm6}. \\
    However, when both $X$ and $V$ are heavy-tailed, we see results consistent with theorem \ref{thm5}.
    In $5$ of $6$ cases when $X$ is lognormal or student-t, $V$ first increases then starts to decline around $N=10^2$ or $10^3$. When $X$ is (t, df=5) and V is (t, df=10), $V$ instead peaks around $N=10^5$ (but declines afterwards). When X is Levy-distributed, utility never goes significantly above zero (optimization completely fails) because the Levy distribution is too heavy-tailed.}
    \label{fig:pythia-random}
\end{figure}

@@

\begin{table}
    \centering
    \caption{Hyperparameters for ACG}
    \label{tableh}
    \begin{tabular}{|c|c|}
        \hline
        Parameter & Value \\
        \hline
        Context length & 133 \\
        \hline
        Iterations & 1000 \\
        \hline
        Candidates per seq. position (k) & 3 \\
        \hline
        Annealing starting value & 9 \\
        \hline
        Annealing ending value & 2 \\
        \hline
    \end{tabular}
\end{table}

>>>>

Because sampling is inefficient at probing the extreme tail, we also find token sequences that optimize Starling 7B-alpha for reward. We considered Greedy Coordinate Gradient (GCG) from \citep{zou2023universal}, a method used to find adversarial suffixes that circumvent jailbreaking, but decided on a faster version of GCG called Accelerated Coordinate Gradient (ACG) from \citep{haizelabs2024acg}. See Table \ref{tableh} for ACG hyperparameters.

@@

Possible distributions for V are normal and t-distribution with df=10.
Possible distributions for X are normal, t with df=3, t with df=5, lognormal, and Levy. (All of these heavy-tailed except for the normal distribution.)
V is scaled to a standard deviation of 2 and X has s.d. of 1 (except for the Levy distribution, which has infinite variance), representing that in ordinary regimes most of the variance comes from utility rather than error.

The results are shown in Figure \ref{fig:best-of-n}. Briefly, the results are consistent with the asymptotic results in theorems \ref{thm5} and \ref{thm6}, showing that overoptimization

\begin{figure}
    \centering
    \makebox[\textwidth][c]{\includegraphics[width=1.5\linewidth]{images/bon_synthetic.png}}
    \caption{When the error $X$ is normal and thus light-tailed, $V$ increases monotonically with $N$, consistent with our Theorem \ref{thm6}. \\
    However, when both $X$ and $V$ are heavy-tailed, we see results consistent with theorem \ref{thm5}. In $5$ of $6$ cases when $X$ is lognormal or student-t, $V$ first increases then starts to decline around $N=10^2$ or $10^3$. When $X$ is (t, df=5) and V is (t, df=10), $V$ instead peaks around $N=10^5$ (but declines afterwards). When X is Levy-distributed, utility never goes significantly above zero (optimization completely fails) because the Levy distribution is too heavy-tailed.}
    \label{fig:best-of-n}
\end{figure}

\subsection{PPO experiment}

In this experiment, we examined PPO with artificially heavy-tailed rewards.

OpenRLHF \citep{hu2024openrlhf} was used to train Pythia 1B with a reward model derived also from Pythia 1B, on the default OpenRLHF prompt dataset.
We used the reward model to represent true utility, and a heavy-tailed error term based on the number of "the" tokens was added to get the proxy reward.
The \texttt{kl\_target=0.5} option was used to dynamically adjust KL penalty, as we mention is done in \citet{ziegler2020finetuning}.
Rewards were not clipped. (Reward clipping can be useful to prevent overoptimization, but is not always used in PPO.)
Response length was limited to 256.

Midway through training, the model generated

@@

\begin{table}
    \centering
    \label{tableh}
    \begin{tabular}{|c|c|}
        \hline
        Parameter & Value \\
        \hline
        Context length & 133 \\
        \hline
        Iterations & 1000 \\
        \hline
        Candidates per seq. position (k) & 3 \\
        \hline
        Annealing starting value & 9 \\
        \hline
        Annealing ending value & 2 \\
        \hline
    \end{tabular} \\
    \caption{Hyperparameters for ACG}
\end{table}
